---
title: "Case Study 1"
subtitle: "AKSTA Statistical Computing"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*The  .Rmd and .html (alternatively the .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Please consider the AI Guidelines as specified in the organizational slides (you can use the table). Points will be deducted if the submitted file is not in a decent form; answers should be answering the concrete questions and not discuss topics in general - please be mindful of our tutors' time and the size of the course. *


\begin{table}[ht]
\centering
\caption{You can use and extend this table to specify your AI usage (if used at all).}
\begin{tabular}{l|rrr}
\toprule
 \textbf{AI usage} & \textbf{Task 1} & \textbf{Task 2} &  \ldots \\
\midrule
\textit{Correcting faulty code}   &  no  &  &    \\
\textit{Improving existing code}   &  little   &  &     \\
\textit{Generating new code}   &  medium   &  &    \\
\textit{Layout and formatting}  &  heavy   &  &     \\
\textit{Text correction}   &    &  &    \\
 \ldots  &    &  &    \\
\bottomrule
\end{tabular}
\end{table}


# 1. Ratio of Fibonacci numbers (1 point)

## a. 
Write two different R functions which return the sequence $r_{i}=F_{i+1}/F_i$
for $i=1,\ldots,n$ where $F_i$ is the $i$th Fibonacci number, once using `for`
and once using `while`.
```{r}
# for version 
fibonacci_ratio_for <- function(n) {
  fib <- c(1, 1)
  if (n >= 2) {
    for (i in 3:(n + 1)) {
      fib[i] <- fib[i-1] + fib[i-2]
    }
  }
  ratios <- fib[2:(n+1)] / fib[1:n]
  return(ratios)
}

# while version
fibonacci_ratio_while <- function(n) {
  fib <- c(1, 1)
  i <- 3
  while (i <= n + 1) {
    fib[i] <- fib[i-1] + fib[i-2]
    i <- i + 1
  }
  ratios <- fib[2:(n+1)] / fib[1:n]
  return(ratios)
}
```


## b. 
Benchmark the two functions for $n=200$ and $n=2000$ (you can use package **microbenchmark** or package **bench** for this purpose). Which function is faster?

```{r}
library(microbenchmark)

# Benchmark for n = 200
benchmark_200 <- microbenchmark(
  for_loop = fibonacci_ratio_for(200),
  while_loop = fibonacci_ratio_while(200),
  times = 100
)

print(benchmark_200)

# Benchmark for n = 2000
benchmark_2000 <- microbenchmark(
  for_loop = fibonacci_ratio_for(2000),
  while_loop = fibonacci_ratio_while(2000),
  times = 100
)

print(benchmark_2000)
```
**Answer**: Based on the benchmarking results shown above, one can conclude that the for loop implementation is slightly faster than the one using the while loop. This makes sense because a while loop needs to evaluate the condition before every iteration, which introduces a slight overhead that accumulates over each iteration.


## c.
Plot the sequence for  $n=100$. 
For which $n$ does it start to stabilize to a value?
What number does the sequence converge to?

```{r}
# compute ratios for n=100
ratios_100 <- fibonacci_ratio_for(100)

# plot ratios
plot(ratios_100, type = "l",
     xlab = "i", ylab = expression(r[i]),
     main = "Ratio of consecutive Fibonacci numbers")

# mark convergence
abline(h = (1 + sqrt(5)) / 2, col = "red", lty = 2)
```
**Answer**: From the plot, one can see that the ratios stabilize at an $i$ value of about 10. The series converges to a value of $\approx 1.618034$ or $\approx 1.618034$ or $\frac{1 + \sqrt{5}}{2}$. This value is famously known as the golden ratio, and it is also a well-known fact that the ratios of the Fibonacci numbers converge to it.



# 2. Gamma function (1 point)

## a. 
Write a function to compute the following for $n$ a positive integer
using the `gamma` function in base R.
$$
\rho_n =\frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n - 2)/2)}
$$

```{r}
rho <- function(n) {
  gamma((n - 1) / 2) / (gamma(1 / 2) * gamma((n - 2) / 2))
}
```


## b. 
Try $n = 2000$. What do you observe? Why do you think the observed behavior happens?
```{r}
rho(2000)
```

**Answer**: We get `NaN`. The reason could be that the function is evaluated with a very large value. After a few experiments with lower values, we observed that anything above 345 produces `NaN`. If we split up the denominator and numerator and calculate them individually, we obtain the following:

```{r formula individual, echo=TRUE}
n_calculation_numerator <- function(n){
  result <- (gamma((n - 1)/2))
  print(result)
}

n_calculation_denominator <- function(n){
  result <- (gamma(1/2) * gamma((n-2)/2))
  print(result)
}

cat("Numerator result for n = 2000: \n")
n_calculation_numerator(2000)
cat("Denominator result for n = 2000: \n")
n_calculation_denominator(2000)
```

Both the numerator and denominator individually return `Inf` for n=2000. The Gamma function grows extremely fast (faster than exponential). Even for moderately large inputs it produces astronomically large numbers. These numbers are so large that they cannot be properly represented by floating-point numbers and R therefore returns `Inf`. Since R cannot divide infinite values by one another and `Inf / Inf` is in general a indeterminate form, the final result is `NaN`.


## c.
Write an implementation which can also deal with large values of $n>1000$. 

**Answer**: The solution to this problem is to handle large values using a log scale. Instead of computing the Gamma function on the standard scale, we compute its logarithm using `lgamma()`. This allows us to avoid computing the extremely large intermediate Gamma values and therefore prevents numerical overflow. After computing the expression on the log scale, we exponentiate the final result to convert it back to the original scale. (ref: https://stackoverflow.com/questions/42821156/how-to-overcome-the-gamma-function-limitation-in-r)

```{r log calculation, echo=TRUE}
rho_stable <- function(n) {
  exp(lgamma((n - 1) / 2) - lgamma(1 / 2) - lgamma((n - 2) / 2))
}
rho_stable(2000)
```


## d. 
Plot $\rho_n/\sqrt{n}$ for different values of $n$. Can you guess the limit of $\rho_n/\sqrt{n}$ for $n \rightarrow\infty$?

```{r}
n_vals <- seq(3, 2000, by = 10)
rho_vals <- sapply(n_vals, rho_stable)

plot(n_vals, rho_vals / sqrt(n_vals), type = "l",
     xlab = "n", ylab = expression(rho[n] / sqrt(n)),
     main = expression(rho[n] / sqrt(n) ~ "as n increases"))


n_vals <- seq(3, 100, by = 10)
rho_vals <- sapply(n_vals, rho_stable)

plot(n_vals, rho_vals / sqrt(n_vals), type = "l",
     xlab = "n", ylab = expression(rho[n] / sqrt(n)),
     main = expression(rho[n] / sqrt(n) ~ "as n increases"))

```
**Answer**: Plotting the ratio for increasing values of $n$ shows that the curve approaches approximately $0.4$. For smaller ranges of $n$, the convergence toward this value appears slower. Based on the behavior for larger $n$, it is reasonable to guess that the limit as $n \to \infty$ is about $0.4$.



# 3. Weak law of large numbers (1.5 points)

The law of large numbers states that if you repeat an experiment independently a large number of times and average the result, what you obtain should be close to the expected value. Formally, the weak law of large numbers states that if 
$X_1, X_2, \ldots X_n$ are independently and identically 
distributed (iid) random variables with a finite expected value
$E(X_i)=\mu<\infty$, then, for any $\epsilon > 0$
$\lim_{n\rightarrow \infty} P(|\bar X_n-\mu|>\epsilon)=0$
where $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$ is the sample mean (the sample mean converges in probability to the expected value).


## a. 

- In R, simulate rolling a fair 6-sided die $n=10\,000$ times using vectorized functions. 
- After each roll, compute the cumulative mean of all previous rolls. 

- Plot the running average against the number of rolls and comment on the result.


```{r}
set.seed(42)
n <- 10000
rolls <- sample(1:6, n, replace = TRUE)
running_mean <- cumsum(rolls) / 1:n

plot(1:n, running_mean, type = "l",
     xlab = "Number of rolls", ylab = "Running mean",
     main = "Running average of a fair die",
     ylim = c(1, 6))
abline(h = 3.5, col = "red", lty = 2)
```

**Answer**: As the number of rolls increases, the running average converges towards the expected value of 3.5. This illustrates the weak law of large numbers, with more observations, our sample mean gets closer to the true population mean.


## b. 
- Replicate the die-rolling experiment 
 $M=50$ times using a for loop. 
This will result in $50$
paths which you can plot against the number of rolls.

- Plot all 50 paths on the same graph to observe the variation in the running averages.

```{r}
M <- 50
n <- 10000
paths <- matrix(NA, nrow = M, ncol = n)

for (m in 1:M) {
  rolls <- sample(1:6, n, replace = TRUE)
  paths[m, ] <- cumsum(rolls) / 1:n
}

plot(1:n, paths[1, ], type = "l", col = rgb(0,0,1,0.3),
     ylim = c(2.5, 4.5),
     xlab = "Number of rolls", ylab = "Running mean",
     main = "50 replications of running average")

# Add the remaining 49 paths
for (m in 2:M) {
  lines(1:n, paths[m, ], col = rgb(0,0,1,0.3))
}

abline(h = 3.5, col = "red", lty = 2)
```

**Answer**: We can see that all paths start with high variability but converge towards the expected value of 3.5 as the number of rolls increases. Some paths approach the expected value from above, others from below, but they all tend to get closer to 3.5 as $n$ increases.


## c.
What is the theoretical expected value of a fair 6-sided die?


**Answer**: For a fair 6-sided die, the possible outcomes (sample space) are $\{1, 2, 3, 4, 5, 6\}$, each with probability $\frac{1}{6}$. The theoretical expected value is:
$$
E(X) = \sum_{i=1}^{6} i \cdot P(X = i) = \sum_{i=1}^{6} i \cdot \frac{1}{6} = \frac{1 + 2 + 3 + 4 + 5 + 6}{6} = \frac{21}{6} = 3.5
$$


## d. 
Comment on how the running mean of the paths behaves relative to the expected value  as $n$ increases.

```{r}
abs_diff <- abs(paths - 3.5)
mean_abs_diff <- colMeans(abs_diff)

plot(1:ncol(paths), mean_abs_diff, type = "l",
     xlab = "Number of rolls", ylab = "Mean absolute difference",
     main = "Convergence to Expected Value")
```

**Answer**: As the number of observations $n$ grows, the average absolute difference between the running mean and the expected value becomes smaller. This behavior is consistent with the weak law of large numbers. At the start, when only a few observations are available, the running mean fluctuates quite a lot because of the randomness in small samples. However, as more die rolls are added, the overall behavior becomes more predictable and centers around the expected value of 3.5.


## e. 

- Modify the code in b. to use a while loop.

- Stop the loop when the proportion of replications where the difference between the running average and the expected value lies outside the interval 
$[-\epsilon,\epsilon]$ is less than 0.05. Use $\epsilon = 0.01$. 

- How high is $n$ at  the end of the loop?

```{r}
set.seed(44)
epsilon <- 0.01
M <- 50
batch <- 100

n_current <- 0
proportion_outside <- 1
running_averages <- matrix(NA, nrow = M, ncol = batch)

while (proportion_outside >= 0.05) {
  n_current <- n_current + batch
  
  if (ncol(running_averages) < n_current) {
    running_averages <- cbind(running_averages, matrix(NA, nrow = M, ncol = batch))
  }
  
  for (m in 1:M) {
    new_rolls <- sample(1:6, batch, replace = TRUE)
    
    if (n_current == batch) {
      running_averages[m, 1:batch] <- cumsum(new_rolls) / (1:batch)
    } else {
      prev_sum <- running_averages[m, n_current - batch] * (n_current - batch)
      new_sums <- prev_sum + cumsum(new_rolls)
      running_averages[m, (n_current - batch + 1):n_current] <-
        new_sums / ((n_current - batch + 1):n_current)
    }
  }
  
  proportion_outside <- mean(abs(running_averages[, n_current] - 3.5) > epsilon)
}

cat("The loop terminated at n =", n_current,
    "with proportion outside epsilon =", proportion_outside, "\n")
```

**Answer**: The loop terminated at $n = 101900$. At this point, fewer than 5% of all replications have a running mean that differs from the expected value (3.5) by more than $\epsilon = 0.01$, meaning that about 95% of all running means lie within this interval.


## f. 

- Repeat the analysis in b, but instead of simulating from a 
discrete uniform distribution, simulate from a continuous standard cauchy distribution using `rcauchy()`. The (standard) cauchy distribution arises from  the ratio of two independent 
mean zero (standard) normally distributed random variables.

- Plot the running mean against increasing 
$n$. Comment on whether the weak Law of Large Numbers seems to hold for this distribution.
Explain why or why not.


```{r}
set.seed(42)
M <- 50
n <- 10000
paths_cauchy <- matrix(NA, nrow = M, ncol = n)

for (m in 1:M) {
  draws <- rcauchy(n)
  paths_cauchy[m, ] <- cumsum(draws) / 1:n
}

plot(1:n, paths_cauchy[1, ], type = "l", col = rgb(0,0,1,0.3),
     ylim = c(-10, 10),
     xlab = "Number of rolls", ylab = "Running mean",
     main = "Running mean - Cauchy distribution")
for (m in 2:M) {
  lines(1:n, paths_cauchy[m, ], col = rgb(0,0,1,0.3))
}
abline(h = 0, col = "red", lty = 2)
plot(1:n, paths_cauchy[1, ], type = "l", col = rgb(0,0,1,0.3),
     ylim = c(-3, 3),
     xlab = "n", ylab = "Running mean",
     main = "Running mean - Cauchy (zoomed)")

for (m in 2:M) {
  lines(1:n, paths_cauchy[m, ], col = rgb(0,0,1,0.3))
}
abline(h = 0, col = "red", lty = 2)
```

**Answer**: The weak Law of Large Numbers does not seem to hold for the Cauchy distribution. This is because the distribution has heavier tails. Those extreme values influence the running mean and since the Cauchy distribution has no finite expected value, the sample mean does not stabilize as the sample size increases.



# 4. Central limit theorem (1.5 points)

The central limit theorem states that, under certain conditions, the sum of a large number of random variables is approximately normal. Consider the case where
$X_1, X_2,\ldots,X_n$ are i.i.d. random variables
with finite expected values ($E(X_i)=\mu<\infty$)
 and variances $(Var(X_i)=\sigma^2<\infty)$.
The central limit theorem states that the cumulative distribution function  (CDF)
of the normalized random variable
$$
Z_n = \frac{\bar X_n - \mu}{\sigma/\sqrt{n}}
$$
converges to the standard normal CDF as $n\rightarrow\infty$.

## a. 
Derive the expected value of the sample mean $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.
 

**Answer**: Placing $\bar{X}_n$ inside the expectation operator and applying linearity, we get:
$$E(\bar{X}_n) = E\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right) = \frac{1}{n}\Big[E(X_1) + E(X_2) + \cdots + E(X_n)\Big]$$
Since all $X_i$ are independently and identically distributed, they share the same mean $E(X_i) = \mu$. Substituting:
$$E(\bar{X}_n) = \frac{1}{n}(\mu + \mu + \cdots + \mu) = \frac{n\mu}{n} = \mu$$
Thus $E(\bar{X}_n) = \mu$, meaning the expected value of the sample mean equals the true mean regardless of sample size.


## b. 
Derive the variance of the sample mean $\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.

**Answer**: Since the sample mean is defined as $\bar{X}_n = \frac{X_1 + X_2 + \ldots + X_n}{n}$, we start with:
$$Var(\bar{X}_n) = Var\left(\frac{X_1 + X_2 + \cdots + X_n}{n}\right)$$
Since the $X_i$ are independent, the variance of their sum equals the sum of their variances. Bringing $\frac{1}{n}$ out of the variance operator (where it gets squared to $\frac{1}{n^2}$):
$$Var(\bar{X}_n) = \frac{1}{n^2} \Big[Var(X_1) + Var(X_2) + \cdots + Var(X_n)\Big]$$
Since all $X_i$ are identically distributed with $Var(X_i) = \sigma^2$:
$$Var(\bar{X}_n) = \frac{1}{n^2}(\sigma^2 + \sigma^2 + \cdots + \sigma^2) = \frac{n\sigma^2}{n^2} = \frac{\sigma^2}{n}$$
Thus $Var(\bar{X}_n) = \frac{\sigma^2}{n}$, meaning the variance of the sample mean decreases as $n$ increases.


## c. 
The central limit theorem makes few assumption on the distribution of the random variables $X_i$.
Consider again the die-rolling experiment where $X_i$
is the random variable giving the number in roll $i$.

- Simulate Die Rolls: For each value of $n$ (where 
$n=\{100,1000,10000\}$), simulate rolling a die 
$n$ times.
Record the mean of the $n$ rolls.

- Repeat the above process $M=100$ times for each value of
$n$. This will give you 100 sample means for each $n$.

- Plot a histogram of the 100 sample means for each value of $n$. Observe the shape of the histograms.

- Discuss how the distribution of the sample means changes as $n$ increases. Comment on whether the histograms appear to approach a normal distribution as $n$, illustrating the Central Limit Theorem.


```{r}
set.seed(44)
M <- 100
n_vals <- c(100, 1000, 10000)

for (n in n_vals) {
  sample_means <- replicate(M, mean(sample(1:6, n, replace = TRUE)))
  hist(sample_means, main = paste("n =", n),
       xlab = "Sample mean", freq = FALSE,
       col = "lightblue",
       xlim = c(3, 4),  
       breaks = 20)
  curve(dnorm(x, mean = 3.5, sd = sqrt(35/12/n)),
        add = TRUE, col = "red", lwd = 2)
}
```

**Answer**: The variance is highest for $n = 100$ and lowest for $n = 10000$. For a fair die, the true mean is $\mu = 3.5$, and as $n$ increases, the sample means spread less around this value.

For $n = 100$, the histogram shows a rough bell shape, still with some irregularity. With $n = 1000$, the distribution becomes smoother and more symmetric. For $n = 10000$, the distribution of sample means closely matches a normal bell shaped curve.

This is consistent with the Central Limit Theorem. As $n$ increases, the distribution of the sample mean becomes increasingly normal and more values concentrate around $\mu = 3.5$. This holds regardless of underlying distribution.


## d. 
Compute expected value $\mu$ and the variance  $\sigma^2$ for the random variable $X_i$ (giving the number in die-roll $i$), which follows a discrete uniform distribution.

```{r}
mu <- (1 + 2 + 3 + 4 + 5 + 6) / 6
cat("Calculated mean: ", mu, "\n")

variance <- ((1^2 + 2^2 + 3^2 + 4^2 + 5^2 + 6^2)/6) - mu^2
cat("Calculated variance: ", variance, "\n")
```


## e.
Now that you computed $\mu$ and $\sigma^2$, 
consider $Z_n$.

- For each value of  $n$ (where $n=\{100,1000,10000\}$)
 and for 
$M=\{100,1000,10000\}$ replications, simulate the rolling of $n$ dice and calculate the $Z_n$.

- Plot histograms of $Z_n$ for each combination of 
$n$ and  $M$.

- Discuss how the distribution of $Z_n$
changes as  $n$  and  $M$ increase.
Comment on whether the distributions of 
$Z_n$ appear to approach a normal distribution, illustrating the Central Limit Theorem.

```{r}
set.seed(44)
mu <- 3.5
sigma2 <- 35/12
n_vals <- c(100, 1000, 10000)
M_vals <- c(100, 1000, 10000)

par(mfrow = c(3, 3), mar = c(4, 4, 2, 1))
for (n in n_vals) {
  for (M in M_vals) {
    Zn <- replicate(M, {
      rolls <- sample(1:6, n, replace = TRUE)
      (mean(rolls) - mu) / sqrt(sigma2 / n)
    })
    hist(Zn, main = paste("n =", n, ", M =", M),
         xlab = expression(Z[n]), freq = FALSE,
         col = "lightblue", xlim = c(-4, 4))
    curve(dnorm(x), add = TRUE, col = "red", lwd = 2)
  }
}
par(mfrow = c(1, 1))
```


    
# 5. Readable and efficient code (1 points)

Read over the code below. 

## a. 
Explain (in text) what the code does.

**Answer**

**(i)** In the first code block, two vectors, x and z, are initialized with 100 observations each from a normal distribution. Even though the variables are uncorrelated, a linear model of x on z is fitted. The residuals from this model are then transformed using the sine function, a small offset is added, and the result is used to update x before fitting the model again. This procedure is repeated four times.
To ensure that x does not become too small during the iterations, a threshold is checked in each step. Finally, the modified vector x is returned.

**(ii)** In the second code block, two variables are again initialized with values from a normal distribution. In this case, however, the variables are correlated. The code then evaluates a linear model of y on x using 4-fold cross-validation. First, the model is fitted using all observations except the first 250. Predictions are then made for these excluded observations, and the RMSE is calculated. This process is repeated three more times, each time leaving out a different block of 250 observations so that every data point serves once as part of the test set.


## b.
Explain (in text) what you would change to make the code more readable.

**Answer**: Both code blocks violate the DRY principle,large parts are repeated and could be replaced by a loop. In particular, we could use a loop for the residual computation and fitting process in (i) and for the train-predict evaluate procedure in (ii). In the loops, hard coded values could be replaced with descriptively named variables, e.g. threshold. Descriptions can then be provided at initialization and where they are updated.


## c.
Change the code according to b. and wrap it in a function. This function should have at most 10 lines (without adding commands to more lines such as `x <- 1; y <- 2`. Such commands will count as 2 lines!). Check that the function called on the same input outputs the same as the provided code.

```{r eval=FALSE}
process_residuals <- function() {
  set.seed(1)
  x <- rnorm(100); z <- rnorm(100)
  for (i in 1:4) {
    if (sum(x >= i/1000) < i) stop(paste("step",i,"requires",i,"obs >=",i/1000))
    x <- i * sin(lm(x ~ z)$residuals) + i * 0.01
  }
  x
}
```

```{r eval=FALSE}
cross_validate <- function() {
  set.seed(1)
  x <- rnorm(1000); y <- 2 + x + rnorm(1000)
  df <- data.frame(x, y)
  r <- numeric(4)
  for (i in 1:4) {
    idx <- ((i-1)*250+1):(i*250)
    r[i] <- sqrt(mean((predict(lm(y~x,data=df[-idx,]),newdata=df[idx,])-df[idx,"y"])^2))
  }
  r
}
```



# 6. Measuring and improving performance (2 points)

Have a look at the code of the function below. It is a function for performing a Kruskal-Wallis test, a robust non-parametric method for testing whether samples come from the same distribution. (Note: we assume no missing values are present in `x`). 
```{r}
kwtest <- function (x, g, ...) 
{
    if (is.list(x)) {
        if (length(x) < 2L) 
            stop("'x' must be a list with at least 2 elements")
        if (!missing(g)) 
            warning("'x' is a list, so ignoring argument 'g'")
        if (!all(sapply(x, is.numeric))) 
            warning("some elements of 'x' are not numeric and will be coerced to numeric")
        k <- length(x)
        l <- lengths(x)
        if (any(l == 0L)) 
            stop("all groups must contain data")
        g <- factor(rep.int(seq_len(k), l))
        x <- unlist(x)
    }
    else {
        if (length(x) != length(g)) 
            stop("'x' and 'g' must have the same length")
        g <- factor(g)
        k <- nlevels(g)
        if (k < 2L) 
            stop("all observations are in the same group")
    }
    n <- length(x)
    if (n < 2L) 
        stop("not enough observations")
    r <- rank(x)
    TIES <- table(x)
    STATISTIC <- sum(tapply(r, g, sum)^2/tapply(r, g, length))
    STATISTIC <- ((12 * STATISTIC/(n * (n + 1)) - 3 * (n + 1))/(1 - 
        sum(TIES^3 - TIES)/(n^3 - n)))
    PARAMETER <- k - 1L
    PVAL <- pchisq(STATISTIC, PARAMETER, lower.tail = FALSE)
    names(STATISTIC) <- "Kruskal-Wallis chi-squared"
    names(PARAMETER) <- "df"
    RVAL <- list(statistic = STATISTIC, parameter = PARAMETER, 
        p.value = PVAL, method = "Kruskal-Wallis rank sum test")
    return(RVAL)
}
```

## a. 
Write a pseudo code outlining what the function does.

**Answer**
```
kwtest <- function (x, g, ...) {
  if (x is a list) {
    if (x contains less than 2 elements)
      abort with message "'x' must be a list with at least 2 elements"

    if (argument g is missing)
      display warning "'x' is a list, so ignoring argument 'g'"

    if (x has non-numeric elements (elements that are not numbers))
      display warning "some elements of 'x' are not numeric and will be coerced to numeric"

    k <- length(x)
    l <- lengths(x)

    if (l contains zeros)
      abort with message "all groups must contain data"

    g <- create empty list
    for i from 1 to k
      repeat value i exactly l[i] times
      append these values to g
    convert g to a categorical variable (factor)

    x <- flatten list x into a single vector
  }
  else
  {
    if (length of x is not equal to length of g)
      abort with message "'x' and 'g' must have the same length"

    convert g to a categorical variable (factor)

    k <- number of distinct levels in g

    if (k is less than 2)
      abort with message "all observations are in the same group"
  }

  n <- get number of elements in x
  if (n is less than 2)
    abort with message "not enough observations"

  r <- assign ranks for each element of x
        (basically assign an order without sorting)
  TIES <- count the frequency of each unique value in x

  for each group in g:
    compute the sum of ranks within that group
    divide that squared sum by the group size
  STATISTIC <- sum of all those values

  STATISTIC <- (12 * STATISTIC / (n * (n + 1)) - 3 * (n + 1)) /
               (1 - sum(TIES^3 - TIES) / (n^3 - n))

  PARAMETER <- k - 1

  PVAL <- upper-tail probability of the chi-squared distribution
          at value STATISTIC with PARAMETER degrees of freedom

  label STATISTIC as "Kruskal-Wallis chi-squared"
  label PARAMETER as "df"

  return a list containing:
    - statistic = STATISTIC
    - parameter = PARAMETER
    - p.value   = PVAL
    - method    = "Kruskal-Wallis rank sum test"
}
```


## b. 
For example data, call the function in two ways: once using `x` as a list and once using `x` as a vector with a corresponding `g` argument. Ensure that the two different function calls return the same thing by aligning the inputs.

```{r}
# x as list
x_list <- list(
  group1 = c(15, 3, 9),
  group2 = c(7, 2, 10),
  group3 = c(1, 8, 4)
)

result_list <- kwtest(x = x_list)


# x as vector with G
x_vec <- unlist(x_list)
g_vec <- factor(rep(seq_along(x_list), lengths(x_list)))

result_vec <- kwtest(x = x_vec, g = g_vec)


# Compare results
cat("--- RESULTS LIST ---\n")
print(result_list)

cat("--- RESULTS VECTOR ---\n")
print(result_vec)
```


## c. 
Make a faster version of `kwtest()` that only computes the Kruskal-Wallis  **test statistic** when the input is a numeric variable `x` and a variable `g` which gives the group membership. You can try simplifying the function above or by coding from the mathematical definition (see https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance). This function should also perform some checks to ensure the correctness of the inputs (use `kwtest()` as inspiration).

```{r}
kwtest_better <- function(x, g) {
  if (!is.numeric(x))
    stop("'x' must be numeric")
  if (length(x) != length(g))
    stop("'x' and 'g' must have the same length")

  g <- factor(g)
  k <- nlevels(g)
  if (k < 2L)
    stop("all observations are in the same group")

  n <- length(x)
  if (n < 2L)
    stop("not enough observations")

  r  <- rank(x)
  ni <- tabulate(g)
  Ri <- vapply(split(r, g), sum, numeric(1))

  H <- (12 / (n * (n + 1))) * sum(Ri^2 / ni) - 3 * (n + 1)

  ti <- tabulate(match(x, sort(unique(x))))
  H  <- H / (1 - sum(ti^3 - ti) / (n^3 - n))

  PARAMETER <- k - 1L
  PVAL      <- pchisq(H, PARAMETER, lower.tail = FALSE)

  names(H)         <- "Kruskal-Wallis chi-squared"
  names(PARAMETER) <- "df"

  list(
    statistic = H,
    parameter = PARAMETER,
    p.value   = PVAL,
    method    = "Kruskal-Wallis rank sum test"
  )
}

# test our new function
kwtest_better(x_vec, g_vec)
```


## d. 
Consider the following scenario. You have samples available from multiple experiments $m=1000$ where you collect the numerical values for the quantity of interest `x` and the group membership for $n$ individuals. The first $20$ individuals in each sample belong to group 1, the following $20$ individuals in each sample belong to group 2, the last $10$ individuals in each sample belong to group 3. Use the following code to simulate such a data structure:
```{r}
set.seed(1234)
m <- 1000 # number of repetitions
n <- 50   # number of individuals
X <- matrix(rt(m * n, df = 10), nrow = m)
grp <- rep(1:3, c(20, 20, 10))
```
  Write a function which performs the Kruskal-Wallis test using the function             `stats:::kruskal.test.default()` for $m$ repeated experiments and returns a vector of $m$ test statistics. The input of this function are a matrix `X` with $m$ rows and $n$ columns and a vector `g` which gives the grouping.


```{r}
kwtest_bulk <- function(X, g) {
  apply(X, 1, function(x) {
    stats:::kruskal.test.default(x, g)$statistic
  })
}

# Run
H <- kwtest_bulk(X, grp)
head(H)
```


## e. 
Write a function which performs the Kruskal-Wallis test using the function in point c. for $m$ repeated experiments and returns a vector of $m$ test statistics.
The input of this function are a matrix `X` with $m$ rows and $n$ columns and a vector `g` which gives the grouping. 

```{r}
kwtest_bulk_2 <- function(X, g) {
  apply(X, 1, function(x) {
    kwtest_better(x, g)$statistic
  })
}

# Run
H <- kwtest_bulk_2(X, grp)
head(H)
```


## f. 
Compare the performance of the two approaches using a benchmarking package on the data generated above. Comment on the results.

```{r}
# Benchmark kwtest
benchmark_kwtest <- microbenchmark(
  kwtest_bulk_lib = kwtest_bulk(X, grp),
  kwtest_bulk_custom = kwtest_bulk_2(X, grp),
  times = 20
)

print(benchmark_kwtest)
```

**Answer**: It appears that the function that uses the library is about three times slower than our custom version. This is quite unexpected because one would expect functions that are implemented directly in R’s standard library to have greater performance. However, it could also be that the library function includes additional logic that our custom function is missing, making it more robust and better suited for functionality that will probably be used by thousands of people worldwide.


## g. 
Now consider vectorizing the function in point c. Compare this approach to the other two.
Comment on the results.

```{R}
kwtest_vectorized <- function(X, g) {
  m <- nrow(X)
  n <- ncol(X)
  g <- factor(g)
  k <- nlevels(g)
  ni <- tabulate(g)

  # Rank all rows at once: m x n matrix
  R <- t(apply(X, 1, rank))

  # Group indicator matrix: n x k (one column per group)
  G <- model.matrix(~ g - 1)

  # Rank sums per group per experiment: m x k
  # Each row of R %*% G gives the k rank sums for that experiment
  Ri <- R %*% G

  # H statistic (vectorized over all m experiments)
  H <- (12 / (n * (n + 1))) * drop(Ri^2 %*% (1 / ni)) - 3 * (n + 1)

  # Tie correction (per row — hard to avoid apply here)
  tie_corr <- apply(X, 1, function(x) {
    ti <- tabulate(match(x, sort(unique(x))))
    sum(ti^3 - ti) / (n^3 - n)
  })

  H / (1 - tie_corr)
}


benchmark_kwtest_with_vec <- microbenchmark(
  kwtest_bulk_lib         = kwtest_bulk(X, grp),
  kwtest_bulk_custom      = kwtest_bulk_2(X, grp),
  kwtest_bulk_vectorized  = kwtest_vectorized(X, grp),
  times = 20L
)

print(benchmark_kwtest_with_vec)
```

**Answer**: As expected the vectorized is much faster than our currently best approach. It is about double as fast as the second best approach.
