---
title: "Case Study 3"
subtitle: "AKSTA Statistical Computing"
author: "Authored by members of Group 88"
date: "2026-05-06"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
---

\begin{table}[ht]
\centering

\begin{tabular}{l|rrrrrr}
\toprule
\textbf{AI usage} & \textbf{Task a} & \textbf{Task b} & \textbf{Task c} & \textbf{Task d} & \textbf{Task e} \\
\midrule
\textit{Correcting faulty code}       & medium & medium & medium & medium & medium \\
\textit{Improving existing code}      & no     & no     & no     & no     & no     \\
\textit{Generating new code}          & no     & no     & no     & no     & no     \\
\textit{Layout and formatting}        & no     & no     & no     & no     & no     \\
\textit{Text correction and rewording}& heavy  & heavy  & heavy  & heavy  & heavy  \\
\bottomrule
\end{tabular}

\begin{tabular}{l|rrrrrr}
\toprule
\textbf{AI usage} & \textbf{Task f} & \textbf{Task g} & \textbf{Task h} & \textbf{Task i} & \textbf{Task j} \\
\midrule
\textit{Correcting faulty code}       & medium & medium & medium & medium & medium \\
\textit{Improving existing code}      & no     & no     & no     & no     & no     \\
\textit{Generating new code}          & no     & no     & no     & no     & no     \\
\textit{Layout and formatting}        & no     & no     & no     & no     & no     \\
\textit{Text correction and rewording}& heavy  & heavy  & heavy  & heavy  & heavy  \\
\bottomrule
\end{tabular}


\begin{tabular}{l|rrrrrr}
\toprule
\textbf{AI usage} & \textbf{Task k} & \textbf{Task l} & \textbf{Task m} & \textbf{Task n} & \textbf{Task o} \\
\midrule
\textit{Correcting faulty code}       & medium & TODO & TODO & TODO & TODO \\
\textit{Improving existing code}      & no & TODO & TODO & TODO & TODO \\
\textit{Generating new code}          & no & TODO & TODO & TODO & TODO \\
\textit{Layout and formatting}        & no & TODO & TODO & TODO & TODO \\
\textit{Text correction and rewording}& heavy & TODO & TODO & TODO & TODO \\
\bottomrule
\end{tabular}

\caption{AI usage for this solution.}
\end{table}


# Data

Load the data set you exported in the final Task of Case Study 2. 
Eliminate all
observations with missing values in the income status variable.

As a reminder, the data set includes world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people entering and leaving a country per 1,000 persons)

for most world entities in 2020. The data was downloaded from https://www.cia.gov/the-world-factbook/about/archives/.
Additional information on continent, subcontinent/region and income status was appended 
to the dataset in Case Study 2.


**Setup**
This setup includes loading the necessary libraries and reading the data from the provided `file_out2.csv`. We drop the unnecessary index column, filter for rows that have no `NA` in `status` and convert `status`, `low_yu` and `high_mnr` to factors.

```{r}
# load libraries
library(readr)
library(dplyr)
library(stringr)
library(readxl)
library(tidyr)
library(ggplot2)

# read data
data = read_csv("file_out2.csv", na=c("", ".")) %>%
  select(-`...1`) %>%
  filter(!is.na(status)) %>%
  mutate(status = factor(status,
                         levels = c("H", "UM", "LM", "L"),
                         ordered = TRUE),
         low_yu = factor(low_yu,
                         levels = c("yes", "no")),
         high_nmr = factor(high_nmr,
                           levels = c("yes", "no")),
         )
# show structure
str(data)
```


# Tasks:

## a. Education expenditure in different income levels

Using **ggplot2**, create a density plot of the education expenditure grouped by income 
status. The densities for the different groups are superimposed in the 
same plot rather than in different plots. Ensure that you order the levels of the 
income status such that in the plots the legend is ordered from High (H) to Low (L).

  * The color of the density lines is black.
  
  * The area under the density curve should be colored differently among the income status levels.
  
  * For the colors, choose a transparency level of 0.5 for better visibility.
  
  * Position the legend at the top center of the plot and give it no title (hint: use `element_blank()`).
  
  * Rename the x axis as "Education expenditure (% of GDP)"

  Comment briefly on the plot.


**Answer**
```{r}
ggplot(data, aes(x = expenditure, fill = status)) +
  geom_density(colour = "black", alpha = 0.5) +
  labs(x = "Education expenditure (% of GDP)", y = "Density") +
  theme(legend.position = "top",
        legend.title = element_blank(),
        legend.justification = "center")
```

The distributions overlap quite strongly, so the differences between income groups are not clearly separated. High-income (H) and upper-middle-income (UM) countries appear to have many observations around roughly 4-5% of GDP and seem somewhat more concentrated than lower-income (L) and lower-middle-income (LM) countries. Lower-income countries are more concentrated at slightly lower education expenditure values, around 2-4% of GDP, but compared to H and UM countries their distribution appears more spread out. The long right tail, especially for the lower-middle-income group, suggests that a few countries spend a much higher share of GDP on education than most others.




## b. Income status in different continents

Investigate how the income status is distributed in the different continents. 

* Using **ggplot2**, create a stacked barplot of absolute frequencies showing how the entities are split into continents and income status. Comment the plot.

* Create another stacked barplot of relative frequencies (height of the bars should be one).
Comment the plot.

* Create a mosaic plot of continents and income status using base R functions. 

* Briefly comment on the differences between the three plots generated to investigate the income 
distribution among the different continents.


**Answer**

```{r}
ggplot(data, aes(x = continent, fill = status)) +
  geom_bar(position="stack") +
  labs(x = "Continent", y = "Count of Countries")
```

The absolute frequency barplot shows that Africa and Asia contain the largest number of countries in the data set, while Oceania has by far the fewest. In Africa most countries fall into the lower-middle (LM) and low (L) income categories, whereas Europe is dominated by high-income (H) countries. The Americas and Asia show a pretty balanced mix across income groups.

```{r}
ggplot(data, aes(x = continent, fill = status)) +
  geom_bar(position="fill") +
  labs(x = "Continent", y = "Share of Countries")
```

Plotting relative frequencies makes the differences in income arrangements more visible and comparable by removing the effect of how many countries each continent contributes. Europe is almost entirely high-income, Oceania and America have a large share of high-income countries as well, while Africa is dominated by low and lower-middle income countries with only a small high-income share. Asia appears comparatively heterogeneous, covering all four income statuses.

```{r}
mosaicplot(table(data$continent, data$status),
           main = "Income Status by Continent",
           xlab = "Continent",
           ylab = "Income Status",
           color = TRUE)
```

The mosaic plot encodes both pieces of information from the previous plots simultaneously: the width of each column reflects how many countries belong to a continent (like the absolute barplot) and the height of each tile reflects the income composition within that continent (like the relative barplot). The absolute barplot is best for reading off counts but hides composition differences for small continents. The relative barplot makes arrangements comparable across continents but loses the size information. The mosaic plot combines both at the cost of being slightly harder to read precise values from.



## c. Income status in different subcontinents

For Oceania, investigate further how the income status distribution is in the different subcontinents.
Use one of the plots in b. for this purpose. Comment on the results.


**Answer**

```{r}
data %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = subcontinent, fill = status)) +
    geom_bar(position = "fill") +
    labs(x = "Subcontinent",
         y = "Proportion of Countries",
         fill = "Income Status") +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))
```

Within Oceania the income status distribution differs strongly between subregions. Australia and New Zealand consist exclusively of high-income countries, while Melanesia and Polynesia are dominated by lower-middle and upper-middle income countries with essentially a moderate amount of high-income entries. It is also worth noting that there are no low-income countries.




## d. Net migration in different continents

* Using **ggplot2**, create parallel boxplots  showing  the distribution of the
net migration rate in the different continents. 

* Prettify the plot (change y-, x-axis labels, etc).

* Identify which country in Asia constitutes the largest negative outlier and 
which country in Asia constitutes the largest positive outlier. 

* Comment on the plot.


**Answer**

```{r}
# with outliers
ggplot(data, aes(x = continent, y = net_migr_rate)) +
  geom_boxplot(fill = "orange", outliers = TRUE, notch = TRUE) +
  labs(title = "Net Migration Rate by Continent (with outliers)",
       x = "Continent", y = "Net Migration Rate")

# without outliers
ggplot(data, aes(x = continent, y = net_migr_rate)) +
  geom_boxplot(fill = "orange", outliers = FALSE, notch = TRUE) +
  labs(title = "Net Migration Rate by Continent (without outliers)",
       x = "Continent", y = "Net Migration Rate")

# identify largest negative and positive outliers in Asia
data %>%
  filter(continent == "Asia") %>%
  filter(net_migr_rate == min(net_migr_rate, na.rm = TRUE)
         | net_migr_rate == max(net_migr_rate, na.rm = TRUE)) %>%
  select(country, net_migr_rate)
```

Most continents have median net migration rates close to zero or slightly negative, suggesting roughly balanced migration or slight net emigration. Europe stands out with a clearly positive median, while Oceania shows the largest spread, ranging from strong net emigration to net immigration. Overall, the plot suggests that continent-level summaries hide substantial differences between individual countries.


## e. Net migration in different subcontinents

The graph in d. clearly does not convey the whole picture. It would be interesting also 
to look at the subcontinents, as it is likely that a lot of migration flows
happen within the continent.

* Investigate the net migration in different subcontinents
using again parallel boxplots.  Group the boxplots by continent (hint: use `facet_grid` with `scales = "free_x"`).

* Remember to prettify the plot (rotate axis labels if needed).

* Describe what you see.


**Answer**

```{r}
# with outliers
ggplot(data, aes(x = subcontinent, y = net_migr_rate)) +
  geom_boxplot(fill = "orange", outliers = TRUE) +
  facet_grid(. ~ continent, scales = "free_x", space = "free_x") +
  labs(title = "Net Migration Rate by Subcontinent (with outliers)",
       x = "Subcontinent", y = "Net Migration Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# without outliers
ggplot(data, aes(x = subcontinent, y = net_migr_rate)) +
  geom_boxplot(fill = "orange", outliers = FALSE) +
  facet_grid(. ~ continent, scales = "free_x", space = "free_x") +
  labs(title = "Net Migration Rate by Subcontinent (without outliers)",
       x = "Subcontinent", y = "Net Migration Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Splitting by subcontinent reveals patterns that were hidden in the continent-level boxplots. In Asia, Western Asia shows the widest spread, ranging from negative to strongly positive net migration rates. In the Americas, Northern America is clearly positive, while Latin America and the Caribbean tends to be negative. African subregions are mostly slightly below zero, suggesting moderate net emigration. European subregions are generally positive or close to zero, especially Northern and Western Europe. Oceania shows strong variation, with Australia and New Zealand positive, while Micronesia and Polynesia are strongly negative.




## f. Median net migration rate per subcontinent.

The plot in task e. shows the distribution of the net migration rate for each 
subcontinent. Here you will work on visualizing only one summary statistic, namely 
the median. 

For each subcontinent, calculate the median net migration rate. Then create a plot which
contains the sub-regions on the y-axis and the median net migration rate on 
the x-axis. 

  * As geoms use points. 
  
  *  Color the points by continent -- use a colorblind friendly palette (see e.g., 
  [here](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)).
  
  * Rename the axes.
  
  * Using `fct_reorder` from the **forcats** package, arrange the levels of subcontinent
  such that in the plot the lowest (bottom) subcontinent contains the lowest median
  net migration
  rate and the upper most region contains the highest median
  net migration
  rate.
  
  * Comment on the plot. E.g., what are the regions with the most influx? 
  What are the regions with the most outflux?

**Answer**

```{r}
library(forcats)

# colorblind friendly palette
cbPalette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
              "#0072B2", "#D55E00", "#CC79A7", "#000000")

# calculate median net migration rate per subcontinent
subcontinent_migr <- data %>%
  group_by(continent, subcontinent) %>%
  summarize(median_nmr = median(net_migr_rate, na.rm = TRUE), .groups = "drop")

# create the plot
ggplot(subcontinent_migr, aes(x = median_nmr, y = fct_reorder(subcontinent, median_nmr), 
  color = continent)) +
  geom_point(size = 3) +
  scale_color_manual(values = cbPalette) +
  labs(x = "Median Net Migration Rate (per 1,000 persons)",
       y = "Subcontinent",
       color = "Continent") +
  theme_minimal()
```
The dot plot shows the median net migration rates for each subcontinent, ordered from the lowest (most outflux) to the highest (most influx). We can see that **Micronesia** and **Polynesia** have the most significant negative median migration rates, indicating substantial outflux. Conversely, **Northern America** and **Western Europe** show the highest positive medians, representing the regions with the most influx. Most other regions cluster around zero, reflecting more balanced migration patterns. Using different colors for continents helps visualize how migration trends often follow geographic groupings.

 
## g. Median youth unemployment rate per subcontinent

For each subcontinent, calculate the median youth unemployment rate. Then create a plot which
contains the sub-regions on the y-axis and the median unemployment rate on 
the x-axis. 

  * Use a black and white theme (`?theme_bw()`)
  
  * As geoms use bars. (hint: pay attention to the statistical transformation taking place
  in `geom_bar()` -- look into  argument `stat="identity"`) 
  
  * Color the bars by continent -- use a colorblind friendly palette.
  
  * Make the bars transparent (use `alpha = 0.7`).
  
  * Rename the axes.
  
  * Using `fct_reorder` from the **forcats** package, arrange the levels of subcontinent
  such that in the plot the lowest (bottom) subcontinent contains the lowest median
  youth unemployment rate and the upper most region contains the highest median
  youth unemployment rate.
  
  * Comment on the plot. E.g., what are the regions with the highest vs lowest youth
  unemployment rate?

**Answer**

```{r}
# calculate median youth unemployment rate per subcontinent
subcontinent_unempl <- data %>%
  group_by(continent, subcontinent) %>%
  summarize(median_yu = median(youth_unempl_rate, na.rm = TRUE), .groups = "drop")

# create the bar plot
ggplot(subcontinent_unempl, aes(x = median_yu, y = fct_reorder(subcontinent, median_yu), 
  fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  labs(x = "Median Youth Unemployment Rate (%)",
       y = "Subcontinent",
       fill = "Continent")
```

The bar plot reveals substantial differences in youth unemployment across subcontinents. **Southern Africa** and **Southern Europe** stand out with the highest median youth unemployment rates, both exceeding 25-30%. On the other end of the spectrum, **South-eastern Asia** and **Central Asia** show the lowest median rates. It is interesting to note that even within continents like Europe, there is a wide gap between the youth unemployment levels in Southern Europe compared to Northern or Western Europe.


## h. Median youth unemployment rate per subcontinent -- with error bars

The value displayed in the barplot in g. is the result of an aggregation, 
so it might be useful to also plot error bars, to have a general idea on
how precise the median unemployment is. This can be achieved by plotting the
error bars which reflect the standard deviation or the interquartile range of 
the variable in each of the subcontinents. 

Repeat the plot from Task g. but include also error bars which reflect the 25% and 75%
quantiles. You can use `geom_errorbar` in **ggplot2**.

**Answer**

```{r}
# calculate median and quantiles
subcontinent_unempl_err <- data %>%
  group_by(continent, subcontinent) %>%
  summarize(median_yu = median(youth_unempl_rate, na.rm = TRUE),
            q25 = quantile(youth_unempl_rate, 0.25, na.rm = TRUE),
            q75 = quantile(youth_unempl_rate, 0.75, na.rm = TRUE),
            .groups = "drop")

# create the bar plot with error bars
ggplot(subcontinent_unempl_err, aes(x = median_yu, y = fct_reorder(subcontinent, median_yu), 
  fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(xmin = q25, xmax = q75), width = 0.4, color = "black") +
  scale_fill_manual(values = cbPalette) +
  theme_bw() +
  labs(x = "Median Youth Unemployment Rate (%) (with IQR error bars)",
       y = "Subcontinent",
       fill = "Continent")
```

Adding error bars representing the interquartile range (IQR) provides a clearer picture of the data's variability. While some regions like **Southern Europe** have a high median and a relatively wide IQR, others like **South-eastern Asia** show much lower variability. This highlights that in certain regions, youth unemployment is consistently high or low, whereas in others, there is a broader range of experiences across different countries.


## i. Relationship between education expenditure and net migration rate

Using **ggplot2**, create a plot showing the relationship between education expenditure and
net migration rate. 

  * Color the geoms based on the income status. 

  * Add a regression line for each development status (using `geom_smooth()`).
  
Comment on the plot. Do you see any relationship between the two variables? Do you see
any difference among the income levels?

**Answer**

```{r}
ggplot(data, aes(x = expenditure, y = net_migr_rate, color = status)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Education Expenditure (% of GDP)",
       y = "Net Migration Rate (per 1,000 persons)",
       color = "Income Status") +
  theme_minimal()
```

The scatterplot displays the relationship between education expenditure and net migration rate, categorized by income status. Overall, there is no strong, universal correlation between the two variables across all groups. However, the regression lines suggest some differences: for high-income (H) countries, the relationship appears slightly positive, while for lower-income groups, the lines are flatter or even slightly negative. This might indicate that in wealthier nations, higher investment in education coincides with higher net immigration, potentially due to broader economic opportunities that attract migrants.


## j. Relationship between youth unemployment and net migration rate

Create a plot as in Task i. but for youth unemployment and net migration rate. Comment briefly.


**Answer**

```{r}
ggplot(data, aes(x = youth_unempl_rate, y = net_migr_rate, color = status)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Youth Unemployment Rate (%)",
       y = "Net Migration Rate (per 1,000 persons)",
       color = "Income Status") +
  theme_minimal()
```

This plot explores the relationship between youth unemployment and net migration. Generally, there is a weak negative relationship for many groups, suggesting that regions with higher youth unemployment may experience slightly higher outflux (lower net migration). This pattern is most visible in the regression lines for lower-middle (LM) and upper-middle (UM) income status countries. For high-income countries, the relationship remains relatively flat, perhaps because migration to these countries is driven by factors other than the immediate youth unemployment rate, such as overall economic stability and higher wages.


## k. Merging population data

Go online and find a data set which contains the 2020 population for the countries of the world
together with ISO codes. 

* Download this data and merge it to the dataset you are working on in
this case study using a left join.
(A possible source: [World Bank](https://data.worldbank.org/indicator/SP.POP.TOTL?end=2020&start=2020)))

* Inspect the data and check whether the join worked well.


**Answer**

```{r}
# read the downloaded population data
pop_data <- read_csv("population_data.csv")

# filter for the year 2020
pop_2020 <- pop_data %>%
  filter(Year == 2020) %>%
  select(ISO3 = `Country Code`, population = Value)

# merge with the main dataset using a left join
data <- data %>%
  left_join(pop_2020, by = "ISO3")

# inspect the join result
summary(data$population)
sum(is.na(data$population))

# show countries where population is missing (if any)
data %>% filter(is.na(population)) %>% select(country, ISO3)
```

The population data for 2020 was successfully downloaded from the World Bank (via DataHub) and merged with our main dataset using the ISO3 codes. After the left join, we performed a quick inspection to verify the results. Most countries now have an associated population value, which will be essential for scaling our analysis in the subsequent tasks. For the few entities where population data might still be missing, we can identify them and decide whether they should be handled as exceptions, though for the majority of recognized countries, the merge worked as expected.


## l. Scatterplot of education expenditure and net migration rate in Europe 

Make a scatterplot of education expenditure and net migration rate for the countries of Europe. 

  * Scale the size of the points according to each country's population. 
 
  * For better visibility, use a transparency of `alpha=0.7`. 
  
  * Remove the legend.
  
  * Comment on the plot.




## m. Interactive plot

On the merged data set from Task k., using function `ggplotly` from package **plotly**  
re-create the scatterplot in Task l., but this time for all countries. Color the points
according to their continent.

When hovering over the points the name of the country, 
the values for education expenditure, net migration rate, and population 
should be shown. (Hint: use the aesthetic `text = Country`. 
In `ggplotly` use the argument `tooltip = c("text", "x", "y", "size")`).

**Hint:** If you are creating a .pdf file, keep in mind that PDF output is static and does not support HTML widgets. 
To avoid build errors, you can wrap interactive plots in a conditional check 
or set the chunk option `eval=FALSE`. Alternatively, you can use the **webshot** 
package to automatically convert interactive elements into static images during the knit process.




## n. Parallel coordinate plot

In **parallel coordinate plots** each observation or data point is depicted as a line 
traversing a series of parallel axes, corresponding to a specific variable or dimension.
It is often used for identifying clusters in the data.

One can create such a plot using the **GGally** R package. You should create 
such a plot where you look at the three main variables in the data set: 
education expenditure, youth unemployment rate and net migration rate. 
Color the lines based on the income status. Briefly comment.




## o. World map visualisation

Create a world map of the education expenditure per country.
You can use the vignette https://cran.r-project.org/web/packages/rworldmap/vignettes/rworldmap.pdf
to find how to do this in R. Alternatively, you can use other packages (such as **ggplot2**, **sf** and **rnaturalearthdata**) to create a map.

