---
title: "Case Study 2"
subtitle: "AKSTA Statistical Computing"
author: "Authored by members of Group 88"
date: "2026-04-27"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!-- *The  .Rmd* **and** *.html (or .pdf) should be uploaded in TUWEL by the deadline. Refrain from using explanatory comments in the R code chunks but write them as text instead. Points will be deducted if the submitted file is not in a decent form.*-->

\begin{table}[ht]
\centering
\caption{You can use and extend this table to specify your AI usage (if used at all).}
\begin{tabular}{l|rrrr}
\toprule
\textbf{AI usage} & \textbf{Task a} & \textbf{Task b} & \textbf{Task c} & \ldots \\
\midrule
\textit{Correcting faulty code}       &  medium   & medium   & medium   &  \ldots \\
\textit{Improving existing code}      &  no       & no       & no       &  \ldots \\
\textit{Generating new code}          &  no       & no       & no       &  \ldots \\
\textit{Layout and formatting}        &  no       & no       & no       &  \ldots \\
\textit{Text correction and rewording}&  heavy    & heavy    & heavy    &  \ldots \\
\ldots  & & & & \\
\bottomrule
\end{tabular}
\end{table}


The CIA World Factbook provides intelligence on various aspects of 266 world entities, including history, people, government, economy, energy, geography, environment, communications, transportation, military, terrorism, and transnational issues. This case study involves analyzing world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people entering and leaving a country per 1,000 persons)

The data was sourced from the [CIA World Factbook Archives](https://www.cia.gov/the-world-factbook/about/archives/). You are required to use `dplyr` for data manipulation, while any package can be used for importing data.

**Setup**
```{r}
library(readr)
library(dplyr)
library(stringr)
library(readxl)
```

# Tasks:

## a. Data Import and Cleaning

Load the following datasets from TUWEL and ensure that missing values are handled correctly and column names are clear. Each dataset should ultimately contain only two columns: **country** and the respective variable. Note that some data sets also contain information on the year when the value was last updated.

  
  * `rawdata_369.txt` which contains the (estimated)  public expenditure on education as a percent of GDP. *Pay attention! The delimiter is 2 or more white spaces (one space would not work as it would separate country names which contain a space); you have to skip the first two lines*. 
  
  * `rawdata_373.csv` which contains the (estimated) youth unemployment rate (15-24) per country
  
  * `rawdata_347.txt`  which contains (estimated) net migration rate per country.


**Answer**
First of all we want to read all the data. According to the task each data set should ultimately contain only two columns: `country` and the respective measured variable, so the rank and the year of the last update are dropped during import.

```{r, message=FALSE}
# read data
raw_data_347 <- read_fwf(
  "rawdata_347.txt",
  col_positions = fwf_widths(c(8, 60, 10, 17),
                             c("rank", "country", "mig_rate", "year")),
  skip = 2,
  na = c("", "NA", "N/A")
) %>%
  select(country, mig_rate)

raw_data_369 <- read_fwf(
  "rawdata_369.txt",
  col_positions = fwf_widths(c(8, 60, 10, 12),
                             c("rank", "country", "edu_exp", "year")),
  skip = 2,
  na = c("", "NA", "N/A")
) %>%
  select(country, edu_exp)

raw_data_373 <- read_csv("rawdata_373.csv") %>%
  rename(country = country_name,
         youth_unempl_rate = youth_unempl_rate)

# show structure of read data
str(raw_data_347)
str(raw_data_369)
str(raw_data_373)
```

The two `.txt` files were parsed with `read_fwf()` from the `readr` package, which can be used to parse fixed-width-files. To do this properly we have to pass the widths of the columns in our files to `read_fwf`. Because the first two rows in our files don't contain any relevant data we use the `skip` parameter to avoid parsing them. The `.csv` file was read with `read_csv()`. Country names in that file contain a long block of trailing whitespaces, which can be safely dropped. This is being handled automatically because the parameter `trim_ws`, which tells our function to trim all leading and trailing whitespaces before parsing, is set to `TRUE` by default. In all three data sets we rename the country column to `country` so that the merging step in part b can use a single, common key.

```{r}
# check for missing values
cat("raw_data_347 contains", sum(is.na(raw_data_347)), "missing values.\n")
cat("raw_data_369 contains", sum(is.na(raw_data_369)), "missing values.\n")
cat("raw_data_373 contains", sum(is.na(raw_data_373)), "missing values.\n")
```

After the import none of the three tibbles contains missing values, so no further NA handling is required at this point. (Missing values will appear naturally in part b when we take the union of all countries through a full join.)




## b. Merging Raw Data

Merge the datasets using `dplyr` on a unique key and retain the union of all observations.

- What key are you using for merging?
- Return the dimensions of the merged dataset.

**Answer**

The only variable that appears in all three raw data sets is the country name, so it is the natural choice for merging. To retain the union of all observations (i.e. every country that appears in at least one of the three files) we use `full_join()`.

```{r}
# merge data
merged_data <- raw_data_347 %>%
  full_join(raw_data_369, by = "country") %>%
  full_join(raw_data_373, by = "country")

# show data, it's structure and dimensions
merged_data
str(merged_data)
dim(merged_data)

# count NA values
apply(is.na(merged_data), MARGIN=2, FUN=sum)
```

The merged dataset has `r nrow(merged_data)` rows (countries) and `r ncol(merged_data)` columns: the country name plus the three measured variables. Countries that are missing from one of the original files now have `NA` for the corresponding variable. For now this is fine, but one has to keep this fact in mind when doing later analyses.




## c. Enriching Data with Income Classification

Obtain country income classification (low, lower-middle, upper-middle, high) from the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519) and merge it with the dataset.

- Identify common variables between datasets. Can they be used for merging? Why or why not?

- Since ISO codes are standardized, download and use the [CIA country data codes](https://www.cia.gov/the-world-factbook/references/country-data-codes/) for merging. Make sure you are not losing any of the countries in your original data set when merging.


**Answer**

The only variable shared between our merged CIA data set and the World Bank income classification is the country name. In principle this column could be used as a merge key, but in practice it is not reliable: the two sources use slightly different spellings and conventions for several countries (for example *Macedonia* vs. *North Macedonia*, *Turkey* vs. *Turkey (Turkiye)* etc.). A name-based join would silently drop every such country. Standardised ISO-3 codes do not have this problem, so we use the CIA *Country Data Codes* file to attach a `GENC` (ISO-3) code to every country in our data and then merge the World Bank classification on that code.

```{r, message=FALSE}
# read country codes from .csv
country_codes <- read_csv("Country Data Codes.csv") %>%
  select(Name, GENC)
```

Before joining we have to make the country names in `merged_data` agree with those in the codes table. A quick anti-join shows which names cannot be matched directly:

```{r}
# check if there are some mismatches
merged_data %>%
  anti_join(country_codes, by = c("country" = "Name")) %>%
  pull(country)
```

Only two countries are affected: *Macedonia* and *Turkey*. After looking them up manually in `country_codes.csv` we found out that this is simple name mismatch. *Macedonia* is called *North Macedonia* in the codes file and *Turkey* is called *Turkey (Turkiye)*. We rename them so that no country is lost in the join.

```{r}
# rename Macedonia and Turkey
merged_data <- merged_data %>%
  mutate(country = recode(country,
                          "Macedonia" = "North Macedonia",
                          "Turkey"    = "Turkey (Turkiye)"))

# merge data with country_codes
merged_data <- merged_data %>%
  left_join(country_codes, by = c("country" = "Name"))

# check if all countries have a country_code
stopifnot(all(!is.na(merged_data$GENC)))

# show data, structure and dimensions
merged_data
str(merged_data)
dim(merged_data)
```

Every country in our data set now carries an ISO-3 code. Next we read the World Bank classification (sheet *List of economies* of the file `CLASS_2025_10_07.xlsx`, which was pulled from the link in the instructions) and keep only the ISO-3 code and the income group. The four income classes are recoded to the short labels `L`, `LM`, `UM`, `H` and stored as an ordered factor so that subsequent tables and plots respect their ordering.

```{r, message=FALSE}
# read income classification data
wb_income_class <- read_excel("CLASS_2025_10_07.xlsx",
                              sheet = "List of economies") %>%
  select(Code, `Income group`) %>%
  rename(GENC = Code, income_group = `Income group`) %>%
  mutate(income_status = recode(income_group,
                                "Low income"          = "L",
                                "Lower middle income" = "LM",
                                "Upper middle income" = "UM",
                                "High income"         = "H"),
         income_status = factor(income_status,
                                levels = c("L", "LM", "UM", "H"),
                                ordered = TRUE)) %>%
  select(GENC, income_status)

# merge incomeclass with our data using country codes
merged_data <- merged_data %>%
  left_join(wb_income_class, by = "GENC")

# show data, structure and dimensions
merged_data
str(merged_data)
dim(merged_data)
```

The use of `left_join()` guarantees that no country from the original data set is dropped. For some countries the World Bank has no classification (a handful of small territories such as *Anguilla* or *Kosovo*) keep `NA` in the new `income_status` column:

```{r}
merged_data %>%
  filter(is.na(income_status)) %>%
  pull(country)
```




## d.  Adding Geographical Information

Introduce continent and subcontinent (or region) data for each country.

- Find and download an appropriate online resource.
- Merge this information into the dataset, naming the final dataset `df_vars`. Make sure you are not losing any of the countries in your original data set when merging.





## e. Data Tidiness and Summary Statistics

- Evaluate the tidiness of `df_vars`
(observational units, variables, fixed vs. measured variables).
Make adjustments to tidy the data, if necessary.

- Create a frequency table for the income status variable and briefly interpret the results.

- Analyze the distribution of income status across continents by computing absolute and relative frequencies. Comment on the findings.

- Using the distribution of income status across continents, identify which countries are the only ones in their income group across the continent. Discuss briefly.





## f. Further Summary Statistics and Insights

- Create a table of average (mean and median) values for expenditure, youth unemployment rate and net migration rate separated into income status. Make sure that in the output, the ordering of the income classes is proper (i.e., L, LM, UM, H or the other way around). Briefly comment the results and any differences between the mean and median. 

- Look at the standard deviation and the interquartile range of the variables per income status instead of the location statistics above. Do you gain additional insights? Briefly comment the results. 


- Extend the analysis of the statistics median and IQR  to **each income status and continent combination**. Play around with displaying the resulting table. Use `pivot_longer()` and/or 
`pivot_wider()` to generate different outputs. 
Discuss the results as well as the readability of the different tables.
 
- Identify countries performing well in terms of both **youth unemployment** and **net migration rate** (top 25% in net migration and bottom 25% in youth unemployment within their continent).





## g. Conditional Probabilities

Estimate the following based on the observed frequencies in the data:

- What is the (posterior or conditional) probability that a European country belongs to the high income group? What is the prior probability that a country belongs to the high income group?

- Given a country has high youth unemployment (above %25), what is the probability that it also has negative net migration?





## h. Simpson’s Paradox Analysis

Investigate whether an overall trend in youth unemployment rate in the high and low income groups reverses when analyzed at the continent level. E.g., does the youth unemployment rate appear lower in low-income countries overall, but higher when controlling for continent? Explain the results and possible reasons behind this paradox.





## i. Data Export

Export the final tidy dataset from e. as a **CSV** with:`;` as a separator; `.` representing missing values; no row names included.  Upload the `.csv` to TUWEL, together with the submission.

